\nonstopmode
\documentclass[a4paper]{article}
\usepackage{natbib}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{amsmath}
%\usepackage{dsfont}
\usepackage{bibentry}

\input{macros}

\newtheorem{theorem}{Theorem}
%\newtheorem*{theorem*}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}

%\newcommand{\qed}{\hfill\ensuremath{\Box}}
% from llncs.cls
%\def\squareforqed{\hbox{\rlap{$\sqcap$}$\sqcup$}}
\def\squareforqed{\ensuremath{\Box}}
\def\qed{\ifmmode\squareforqed\else{\unskip\nobreak\hfil
\penalty50\hskip1em\null\nobreak\hfil\squareforqed
\parfillskip=0pt\finalhyphendemerits=0\endgraf}\fi}

\newenvironment{proof}[1][]{\noindent\ifthenelse{\equal{#1}{}}{{\it
      Proof.}}{{\it Proof #1.}}\hspace{2ex}}{\qed\bigskip}
\newenvironment{proof*}[1][]{\noindent\ifthenelse{\equal{#1}{}}{{\it
      Proof.}}{{\it Proof #1.}}\hspace{2ex}}{\bigskip}

\newcommand{\wk}{\mathord{\uparrow}}
\newcommand{\WW}{\mathbb{W}}
\newcommand{\tAr}{\mathsf{Ar}}
\newcommand{\Ar}[1]{\tAr(#1)}

\title{Normalization by Evaluation for \\ Intuitionistic Propositional Logic}
\author{Andreas Abel}
\date{July 2018}

\begin{document}
\maketitle

\nobibliography{auto-nbeSum}

\section{Intuitionistic Propositional Logic (IPL)}

\para{Formulas} and hypotheses lists (contexts).
\[
\begin{array}{lllrll}
  \Atom & \ni & P,Q & & & \mbox{atomic propositions} \\
  \Form & \ni & A,B,C,\dots
     & ::= & \top \mid A \land B \mid A \imp B
       & \mbox{negative formulas} \\
     &&& \mid & \bot \mid A \lor B \mid P
       & \mbox{positive formulas} \\
\\[1ex]
  \Cxt & \ni & \Gamma,\Delta & ::= & \cempty & \mbox{empty context}
\\ &&& \mid & \Gamma.A & \mbox{context extension}
\end{array}
\]
We write just $A$ for the singleton context $\cempty.A$.

Indices $x$ into the context $\Gamma$ of hypothesis, locating hypothesis
$A$, are written $x : \Hyp[\Gamma]A$ and defined inductively by the
following constructors:
\[
  \ru{}{\ttop : \Hyp[\Gamma.A]A}
\qquad
  \ru{x : \Hyp[\Gamma]A
    }{\pop[B] x : \Hyp[\Gamma.B]A}
\]
Deriviations $t$ of the truth of a formula $A$ under assumptions
$\Gamma$ (judgement $\Gamma \der A$)
are written $t : \Tm[\Gamma]A$ and defined inductively as follows:

\paradot{Implication and hypotheses}
\begin{gather*}
  \ru{x : \Hyp[\Gamma]A}{\hyp x : \TmG A}
\qquad
  \ru{t : \Tm[\Gamma.A]B}{\impI t : \TmG{A \imp B}}
\qquad
  \ru{t : \TmG{A \imp B} \qquad
      u : \TmG A
    }{\impE t u : \TmG B}
\end{gather*}

\paradot{Truth and conjunction}
\begin{gather*}
  \ru{}{\trueI : \TmG\top}
\qquad
  \ru{t : \TmG A \qquad
      u : \TmG B
    }{\andI t u : \TmG{A \land B}
    }
\\[1ex]
  \ru{t : \TmG{A \land B}
    }{\andE1 t : \TmG A
    }
\qquad
  \ru{t : \TmG{A \land B}
    }{\andE2 t : \TmG B
    }
\end{gather*}

\paradot{Absurdity and disjunction}
\begin{gather*}
  \ru{t : \TmG A
    }{\orI1 t : \TmG {A \lor B}
    }
\qquad
  \ru{t : \TmG B
    }{\orI2 t : \TmG {A \lor B}
    }
\\[1ex]
  \ru{t : \TmG \bot
    }{\falseE t : \TmG C}
\qquad
  \ru{t : \TmG {A \lor B} \qquad
      u : \Tm[\Gamma.A]C  \qquad
      v : \Tm[\Gamma.B]C
    }{\orE t u v : \TmG C}
\end{gather*}

\section{Subformula property and normal derivations}

For every derivable judgement $\Gamma \der A$ there are infinitely
many derivations $t : \TmG A$, since arbitrary detours are allowed.
For instance to prove $\top$ we can also proceed very indirectly by
introducing a hypothesis $\top$ and then eliminating it by proof
$\trueI$:
\[\impEP{\impIP{\hyp{\ttop}}}{\trueI} : \Tm[\cempty]\top.\]
A sensible restriction is that the proof rules guarantee that only
\emph{subformulas} of $A$ and the hypotheses in $\Gamma$ are mentioned when
deriving $\Gamma \der A$.

The following indexed grammar for \emph{normal} derivations $\NfG A$ of
judgement $\Gamma \der A$ ensures the subformula property.  It is
defined mutually with a grammar $\NeG A$ of \emph{neutral} derivations.
Neutral derivations are ``straightline'' consequences of hypotheses
(without case distinction or absurdity elimination).

\paradot{Implication and hypotheses}
\begin{gather*}
  \ru{x : \Hyp[\Gamma]A}{\hyp x : \NeG A}
\qquad
  \ru{t : \Nf[\Gamma.A]B}{\impI t : \NfG{A \imp B}}
\qquad
  \ru{t : \NeG{A \imp B} \qquad
      u : \NfG A
    }{\impE t u : \NeG B}
\end{gather*}

\paradot{Truth and conjunction}
\begin{gather*}
  \ru{}{\trueI : \NfG\top}
\qquad
  \ru{t : \NfG A \qquad
      u : \NfG B
    }{\andI t u : \NfG{A \land B}
    }
\\[1ex]
  \ru{t : \NeG{A \land B}
    }{\andE1 t : \NeG A
    }
\qquad
  \ru{t : \NeG{A \land B}
    }{\andE2 t : \NeG B
    }
\end{gather*}

\paradot{Absurdity and disjunction}
\begin{gather*}
  \ru{t : \NfG A
    }{\orI1 t : \NfG {A \lor B}
    }
\qquad
  \ru{t : \NfG B
    }{\orI2 t : \NfG {A \lor B}
    }
\\[1ex]
  \ru{t : \NeG \bot
    }{\falseE t : \NfG C}
\qquad
  \ru{t : \NeG {A \lor B} \qquad
      u : \Nf[\Gamma.A]C  \qquad
      v : \Nf[\Gamma.B]C
    }{\orE t u v : \NfG C}
\end{gather*}

\paradot{Embedding}
\[
  \ru{t : \NeG P
    }{\neu t : \NfG P}
\]

The introduction rules generate normal forms ($\tNf$s) from $\tNf$s.  The elimination
rules for negative formulas generate neutrals ($\tNe$s) from $\tNe$s, where
$\timpE$ requires a side argument in normal form.

Some care is needed with  the elimination rules for
positive formulas which are $\tfalseE$ and $\torE$.
These prove an arbitrary formula $C$, which, for the sake of the
subformula property, should not be subject to
further elimination but directly prove the goal.
Thus, in addition to restriction the elimination to neutrals
(perform case distinction only on neutrals),
we require them to produce a normal form
in case of $\torE$ from normal branches.

Any neutral derivation at atomic proposition $P$ is considered normal.
The restriction on atoms is not needed for the subformula property,
but forces derivations to be $\eta$-long.
We could drop the restriction, but even with this restriction, the
calculus is complete, meaning that for every derivation ($\tTm$) there
exists a normal derivation ($\tNf$).  This statement is called
\emph{normalization} and we will prove it constructively in the
following.


\section{Categories and Presheaves}

We define order-preserving embeddings (OPEs) $\tau : (\Delta \leq \Gamma)$ of a
context $\Gamma$ into a larger context $\Delta$ inductively by the
following constructors.
\begin{gather*}
  \ru{}{\id[\Gamma] : \Gamma \leq \Gamma}
\qquad
  \ru{\tau : \Delta \leq \Gamma
    }{\weak[A] \tau : \Delta.A \leq \Gamma}
\qquad
  \ru{\tau : \Delta \leq \Gamma
    }{\lift[A] \tau : \Delta.A \leq \Gamma.A}
\end{gather*}
For example $\weak [D] (\lift[C] (\weak[B] \id[A])) : A.B.C.D \leq
A.C$.  OPEs let us introduce extra, unused hypotheses into the
context.  Derivations of $\Gamma \der A$ can be weakened to
derivations $\Delta \der A$ for $\Delta \leq \Gamma$.

OPEs are closed under composition:  If $\tau : \Delta \leq \Gamma$ and
$\tau' : \Phi \leq \Delta$ then $\tau \comp \tau' : \Phi \leq \Delta$.
Let $\OPE$ be the category of contexts (as objects) and OPEs (as morphisms).
\begin{exercise}
  Define composition of OPEs by induction and show that $\OPE$ is
  indeed a category.
\end{exercise}

The context-indexed sets $\Hyp A$, $\Tm A$, $\Nf A$, and $\Ne A$ are
closed under weakening. \Eg{} $\tHyp$: for each $\tau : \Delta \leq \Gamma$
we have a function
\[
  \Hyp[\tau]A : \Hyp[\Gamma] A \to \Hyp[\Delta] A,
\]
\ie, a morphism in the category $\SET$ of sets and functions.
The collection $\Hyp A : (\tau : \Delta \leq \Gamma) \to
(\Hyp[\Gamma]A \to \Hyp[\Delta]A)$ of these function constitutes a
contravariant functor $\Hyp A : \OPE \to \SET$ mapping OPEs to
functions.  In other words, $\Hyp A$ is a \emph{presheaf} over $\OPE$
for each $A$, and so are $\Tm A$, $\Nf A$, and $\Ne A$.

\begin{exercise}
  Prove that $\Hyp A$, $\Tm A$, $\Nf A$, and $\Ne A$ are presheaves
  over $\OPE$ by defining the map functions and proving the functor laws!
\end{exercise}

A function $f$ between presheaves $\A$ and $\B$ is defined pointwise;
we write $f : \A \todot \B$ for the $\Cxt$-indexed collection
\[
  f_\Gamma : \A_\Gamma \to \B_\Gamma
\]
of functions (morphisms in $\SET$). Presheaves and functions between
them make a category by virtue of pointwise identity and composition.
However, typically the functions we consider are \emph{natural} in the
context index, \ie, they are \emph{natural transformations} between
presheaves, which means that they commute with OPEs.  Given $\tau :
\Delta \leq \Gamma$ and $f : \A \todot \B$, naturality means
\[
  f_\Delta \comp \A_\tau = \B_\tau \comp f_\Gamma .
\]
In words, first weakening with $\tau$ and then applying $f$ has the
same effect as applying $f$ first and perform the weakening later.
This guarantees some form of \emph{parametricity} of $f$ in the
context, in particular, $f$ cannot make decisions based on the length
of the context or its precise contents.

Presheaves over $\OPE$ and natural transformations form a category
$\PSh$, traditionally called $\widehat{\OPE}$.

\begin{exercise}
  Prove that $\PSh$ is indeed a category.
\end{exercise}

\begin{exercise}
  Prove that the following functions are morphisms in $\PSh$:
\[
\begin{array}{lllll}
  \thyp & : & \Hyp A & \todot & \Ne A \\
  \tandE1 & : & \Ne{A \land B} & \todot & \Ne A \\
  \torI1  & : & \Nf A & \todot & \Nf{A \lor B} \\
  \tfalseE & : & \Ne \bot & \todot & \Nf C \\
\end{array}
\]
  Find other examples of presheaf morphisms.
\end{exercise}


\section{NbE for the negative fragment}

We wish to define a normalization function for derivations of any
formula $A$
\[
  \tnorm^A : \Tm A \todot \Nf A .
\]
Normalization by evaluation (NbE) achieves normalization by first
evaluating terms, obtaining values and functions that compute by means
of the computation in the meta language, i.e., in $\SET$.  As a second
step, these values are reified back into syntax, yielding a normal
form.  Technically, we define a suitable presheaf $\den A$ for each
formula $A$ and two functions
\[
\begin{array}{lllll}
  \teval^A & : & \Tm A  & \todot & \den A \\
  \treify^A  & : & \den A & \todot & \Nf A \\
\end{array}
\]
which compose to $\tnorm$.

\subsection{Semantics, reflection and reification}

For now, let us focus on the negative
formulas $\top$, $A \land B$, $A \imp B$ and atoms $P$.
\[
\begin{array}{lll@{\qquad}lll}
  \den P & = & \Nf P
    & \reify[\Gamma] P t & = & t \\
  \den \top & = & \hatone
    & \reify[\Gamma] \top \_ & = & \trueI \\
  \den{A \land B} & = & \den A \hattimes \den B
    & \reify[\Gamma]{A \land B} {(a,b)} & = & \andIPP {\reify[\Gamma] A a} {\reify[\Gamma] B b} \\
\end{array}
\]
Herein, we use the unit presheaf $\hatone_\Gamma = 1$ where $1$ is the
unit set, and the pointwise product of presheaves
$(\A \hattimes \B)_\Gamma = \A_\Gamma \times \B_\Gamma$.  We note that
by these definitions, the category $\PSh$ has products, with the
pointwise terminal morphisms:
\[
\begin{array}{lll@{\qquad}lll}
   ! & : & \A \todot \hatone &
     [\_,\_] & : & (f : \C \todot \A) (g : \C \todot \B) \to (\C
                   \todot \A \hattimes \B) \\
   !_\Gamma(a) & = & ()       &
     [f,g]_\Gamma(c) & = & (f(c),g(c)) \\
\end{array}
\]


The interpretation $\den{A \imp B}$ of implication will follow the
Brouwer-Heyting-Kolmogorov (BHK) interpretation of intuitionistic
logic:
\begin{center}\itshape
  A proof of $A \imp B$ is a method turning any proof of $A$ into a
  proof of $B$.
\end{center}
However, the direct lifting of the functions space $\den{A \imp
  B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
not give a presheaf because of the contravariant occurrence of
presheaf $\den A$.  Instead, we force monotonicity by quantifying over
all OPEs of $\Gamma$:
\[
  \den{A \imp B}_\Gamma = \prod_{\Delta : \Cxt} \prod_{\tau : \Delta \leq \Gamma} (\den A
  _\Delta \to \den B _\Delta)
\]
Reification is also challenging.  Let $f : \den{A \imp B}_\Gamma$ and
attempt
%$\tau : \Delta \leq \Gamma$
\[
  \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
    = \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
\]
where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
which intuitively should represent the new hypothesis $\top :
\Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural
transformation $\Hyp A \todot \den A$ which \emph{reflects} variables
into the semantics.  In order to define it by
induction on $A$, we need to generalize it to
\[
  \treflect^A : \Ne A \todot \den A
  .
\]
Let us make sure we can define it at atoms and conjunctions:
\[
\begin{array}{lllll}
  \treflect^P & = & \tne & : & \Ne P \todot \Nf P \\
  \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
  \treflect^{A \land B} & = & \langle\, \treflect^A \comp \andE1  , \;
                             \treflect^B \comp \andE2 \rangle
    & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
\end{array}
\]
Observe the eliminations $\andE i$ introduced to reflect conjunctive hypotheses!

Now, we can complete the story for implication.  Let $\tau : \Delta
\leq \Gamma$ and $a : \den A_\Delta$ and set:
\[
\begin{array}{lll}
  \treflect^{A \imp B} & : & \Ne{A \imp B} \todot \den{A \imp B} \\
  \reflect[\Gamma]{A \imp B} t\;\Delta\;\tau\;a & = &
    \reflect[\Delta]B{(\impEP {\Ne[\tau] B t} {(\reify[\Delta] A a)})}
\\[1ex]
  \treify^{A \imp B} & : & \den{A \imp B} \todot \Nf{A \imp B} \\
  \reify[\Gamma]{A \imp B} f & = &
    \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;(\reflect[\Gamma.A]A{(\hyp\ttop)})}
\end{array}
\]

\begin{exercise}
  Let $A = ((P \imp Q) \land (Q \imp \top)) \imp Q$ for some atoms
  $P,Q$.  Compute $\reflect[A]A{(\hyp \ttop)}$.
\end{exercise}

\begin{exercise}
  Let $A = (P \imp Q) \imp (P \imp Q)$ for some atoms $P,Q$.
  Compute $\treify^A \comp \treflect^A$.
\end{exercise}

\begin{exercise}
  Using the constructions in this section, show that the category of
  presheaves and natural transformations is cartesian closed.
\end{exercise}


\subsection{Evaluation}

Evaluation is turning an expression $t : \TmG A$ into a value in
$\den A$.  However, a direct definition $\teval^A : \Tm A \todot \den A$
by induction on the expression fails in case $\timpI$:
\[
\begin{array}{lll}
   \teval^{A \imp B}_\Gamma & : & \TmG A \to \den {A \imp B} _\Gamma \\
   \teval^{A \imp B}\,(\impI t)\;\Delta\;a & = & ? \\
\end{array}
\]
To be able to pass the argument $a$ to the function $\teval\,(\impI
t)$, we have to generalize evaluation to take an valuation of the
context into account.
\[
  \teval^{A \imp B}_\Gamma  : \den \Gamma \todot \den {A \imp B}
\]
Here, $\den \Gamma$ is the presheaf obtained as the product of the
hypotheses contained in $\Gamma$:
\[
\begin{array}{lll}
  \den{\cempty} & = & \hatone \\
  \den{\Gamma.A} & = & \den\Gamma \hattimes \den A \\
\end{array}
\]
Looking up values in the environment is the following family of
natural transformations:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 2 l {\tlookup^A_\Gamma} & : & \HypG A \to (\den \Gamma \todot \den A) \\
  \lookup[\Gamma.A]A \ttop    & \Delta\,(\gamma,a) & = & a \\
  \lookup[\Gamma.B]A {\pop x} &\Delta\,(\gamma,a) & = & \lookup[\Gamma]A x\, \Delta\, \gamma \\
\end{array}
\]
\begin{exercise}
  Express $\tlookup$ via the projections
  $\pi_i : \A_1 \hattimes \A_2 \todot \A_i$
  out of the presheaf product.
\end{exercise}
We are ready to define evaluation:
\[
\begin{array}{lcl}
  \evalG A {\hyp x} & = & \lookupG A x \\
  \evalG {A \imp B} {\impI t}\,\Delta\,\gamma\,\Delta'\,(\tau : \Delta' \leq \Delta)\,a
    & = & \eval[\Gamma.A]B t\,\Delta'\,(\den\Gamma_\tau\gamma,a) \\
  \evalG {B} {\impE t u}\,\Delta\,\gamma & = &
    \evalG {A \imp B} t\,\Delta\,\gamma\,\Delta\,\id[\Delta]\,
      (\evalG A u\,\Delta\,\gamma) \\
\end{array}
\]
\begin{exercise}
  Complete the definition of $\teval$ for conjunctions.
\end{exercise}
\begin{exercise}
  Express $\teval$ in terms of the operations of the CCC of
  presheaves!
\end{exercise}

\subsection{Normalization}

Normalization is reification after evaluation in the identity
environment:
\[
\begin{array}{lcl}
\tnorm^A & : & \Tm A \todot \Nf A \\
\norm[\Gamma] A t & = & \reify[\Gamma]A {(\eval[\Gamma] A t \,\Gamma\, \env[\Gamma])}
\end{array}
\]
The identity environment $\env[\Gamma]$ is defined by induction on
$\Gamma$ as follows:
\[
\begin{array}{lcl}
  \env[\Gamma] & : & \den \Gamma _\Gamma \\
  \env[\cempty] & = & () \\
  \env[\Gamma.A] & = & (\den\Gamma_{\weak[A] \id[\Gamma]}(\env[\Gamma]), \;
                          \reflect[\Gamma.A] A {(\hyp \ttop)})
\end{array}
\]

\subsection{Digression: The CCC of reifiable presheaves}

Instead of defining the semantics $\den A$ of formula $A$ by induction
on $A$, we can consider a category whose objects are interpretations
of formulas, and show that this category is cartesian closed.

An object in this category would be a quadruple consisting of:
\[
\begin{array}{lll@{\qquad}l}
  \A & : & \PSh  & \mbox{a presheaf} \\
  A  & : & \Form & \mbox{a formula} \\
  \treflect^A & : & \Ne A \todot \A
    & \mbox{a method to reflect neutrals into $\A$} \\
  \treify^A & : & \A \todot \Nf A
    & \mbox{a method to reify values of $\A$} \\
\end{array}
\]
\begin{exercise}
  Complete this definition by a suitable notion of morphism.
\end{exercise}
\begin{exercise}
  Show that this category is cartesian closed.
\end{exercise}


%\section{Extension to the positive connectives}
\section{Disjunction and absurdity}

For sets $S_1,S_2 : \SET$ let $S_1 + S_2$ denote the disjoint union with
injections $\inj{i} : S_i \to S_1 + S_2$ and case distinction
distinction $[f_1,f_2] : S_1 + S_2 \to T$ for $f_i : S_i \to T$
such that $[f_1,f_2] \comp \inj i = f_i$.

Presheaf coproduct is given pointwise by $(\A \hatplus \B)_\Gamma = \A_\Gamma + \B_\Gamma$.
\begin{exercise}
  Show that $\hatplus$ is indeed a coproduct in $\PSh$.
\end{exercise}


\subsection{Semantics, reflection and reification}

The first idea how to model disjunction,
$\den{A \lor B}_\Gamma = \den{A}_\Gamma + \den{B}_\Gamma$, fails
because we cannot reflect disjunctive hypotheses.  \Eg, an element of
$\den{A \lor B}_{A \lor B}$ should be of the form $\inj1\,a$ or
$\inj2\,b$, but the decision on first or second injection cannot be
made here;  it depends on the hypothesis $A \lor B$.  The solution is
to extend the semantics to allow case distinction on variables (and
neutrals) before making the decisions needed for disjunctions.

Given a presheaf $\A$, we define a new presheaf $\Cover \A$
inductively by the following constructors:
\begin{gather*}
  \ru{a : \A_\Gamma}{\returnC a : \Cover\A_\Gamma}
\qquad
  \ru{t : \NeG \bot}{\falseC t : \Cover\A_\Gamma}
\\[1ex]
  \ru{t : \NeG{C \lor D} \qquad
      c : \Cover\A_{\Gamma.C} \qquad
      d : \Cover\A_{\Gamma.D}
    }{\orC t c d : \Cover\A_\Gamma}
\end{gather*}
\begin{exercise}
  Show that $\Cover\A$ is indeed a presheaf.
\end{exercise}
We can think of $c : \Cover\A_\Gamma$ as a decision tree with leaves in $\A$
and nodes labelled by terms we case on.  Nodes of type $\tfalseC$ have
no subtrees since \emph{ex falsum quod libet}, and $\torC$ nodes
casing on $t : \NeG{C \lor D}$ have
two subtrees, one having a new hypothesis $C$, and one having a new
hypothesis $C$.

We can replace all the leaves according to a function $f : \A \todot \B$,
this makes $\tCover$ a \emph{functor} in the category of presheaves.  We can
also replace all the leaves by new case trees via a natural transformation
\[
   \tjoinC ~:~ \Cover{\Cover\A} \todot \Cover\A
\]
which makes $\tCover$ a \emph{monad} in $\PSh$.
\begin{exercise}
  Define the functorial action $\Cover f$ and $\tjoinC$ and prove the
  monad laws.
\end{exercise}
Further, if the leaves of a case tree are normal forms, we can turn it into one big
normal form via:
\[
  \tpasteNf ~:~ \Cover{\Nf A} \todot \Nf A
\]
\begin{exercise}
  Define $\tpasteNf$.
\end{exercise}

\noindent
We have everything now to model disjunction:
\[
\begin{array}{lcl}
\den{A \lor B} & = & \Cover{\den A \hatplus \den B}
\\[2ex]
\treflect^{A \lor B} & : & \Ne{A \lor B} \todot \den{A \lor B} \\
\reflect[\Gamma]{A \lor B}t & = & \orC t
  {(\returnCP {\injP1 {\reflect[\Gamma.A]A{(\hyp \ttop)}}})}
  {\\ && \qquad\ (\returnCP {\injP2 {\reflect[\Gamma.B]B{(\hyp \ttop)}}})}
\\[2ex]
\treifyOr^{A \lor B} & : & \den{A} \hatplus \den{B} \todot \Nf{A \lor B} \\
\reifyOr[\Gamma]{A \lor B}(\inj1\, a) & = & \orIP1 {\reify[\Gamma]A a} \\
\reifyOr[\Gamma]{A \lor B}(\inj2\, b) & = & \orIP2 {\reify[\Gamma]B b}
\\[1.5ex]
\treify^{A \lor B} & : & \den{A \lor B} \todot \Nf{A \lor B} \\
\treify^{A \lor B} & = & \tpasteNf \comp \Cover{\treifyOr^{A \lor B}} \\
\end{array}
\]
Reification works by first reifying the leaves of the case tree using
$\treifyOr$, and then turning the whole tree into a normal form using $\tpasteNf$.
\begin{exercise}
  Compute $\reify[P \lor Q]{P \lor Q} (\reflect[P \lor Q]{P \lor Q}
  (\hyp \ttop))$.
\end{exercise}
For absurdity, we use the presheaf $\hat0_\Gamma = \{\}$ with initial
morphism $\botelim : \hat0 \todot \C$.
\[
\begin{array}{lcl}
\den{\bot} & = & \tCover\,{\hat0}
\\[2ex]
\treflect^{\bot} & : & \Ne{\bot} \todot \den{\bot} \\
\reflect[\Gamma]{\bot}t & = & \falseC t
\\[2ex]
\treify^{\bot} & : & \den{\bot} \todot \Nf{\bot} \\
\treify^{\bot} & = & \tpasteNf \comp \Cover{\botelim} \\
\end{array}
\]
To reify a value $c : \den\bot$, observe that because $\hat0$ is
empty, $c$ can only be a case tree without leaves, \ie, all branches
end in $\tfalseC$ nodes.  Thus, the mapping
$\Cover{\botelim} : \Cover{\hat0} \todot \Cover{\Nf\bot}$ is merely a
type cast.  The subsequent $\tpasteNf$ will turn the leafless tree $c$
into a normal form consisting entirely of splits $\torE$ and $\tfalseE$.

\begin{exercise}
Prove that $\treify^{A} \comp \treflect^{A}$ produces the long
$\eta$-normal form if applied to variables (\ie, hypotheses).
\end{exercise}



\subsection{Evaluation}

Evaluation poses one more challenge. Consider the term:
\[
\ru{x : \HypG {C \lor D} \qquad
    f : \Tm[\Gamma.C]{A \imp B} \qquad
    g : \Tm[\Gamma.D.A]{B} \qquad
    a : \TmG A
  }{\impEP{\orE {(\hyp x)} f {(\impI g)}} a : \TmG B}
\]
There is a redex $\impEP {\impI g} a$ separated by a $\torE$.  The latter
would naturally evaluate to a case tree with functions in the
leaves.  To apply it to the evaluation of $a$, we have to apply each
leaf to that argument.  We facilitate this by implementing
\[
  \tpaste^{A \imp B} : \Cover{\den{A \imp B}} \todot \den{A \imp B}
\]
that turns a case tree of functions into a function, which expects an
argument that will then be passed to all the functions in the case
tree.  In fact, we need pasting at all formulas $A$:
\[
\begin{array}{lcl}
\tpaste^A & : & \tCover\,{\den A} \todot \den A \\[1ex]
\tpaste^P & = & \tpasteNf \\
\tpaste^\bot & = & \tjoinC \\
\tpaste^{A \lor B} & = & \tjoinC \\
\tpaste^{A \land B} & = & \langle\,
  \tpaste^A \comp \Cover{\pi_1} \, , \;
  \tpaste^B \comp \Cover{\pi_2} \,\rangle \\[1ex]
\multicolumn 3 {l} {\paste[\Gamma]{A \imp B} c \;
  \Delta \; (\tau : \Delta{\leq}\Gamma) \; (a : \den A _\Delta) \, = \,
  (\tpaste^B \comp \mapC \phi \comp \Cover{\den {A \imp B}_\tau}\, c)
}
\end{array}
\]
Here, we need a stronger version of functoriality of $\tCover$ which
only requires an
\[
  \phi : \prod_{\Phi} \prod_{\delta : \Phi \leq \Delta} \A_\Phi
             \to \B_\Phi
\]
to produce a $\tmapC_\Delta t\; \phi : \Cover\A_\Delta \to \Cover\B_\Delta$.
In our case, $\A = \den{A \imp B}$ and $\B = \den{B}$ and
\[
  \phi\;\Phi\;\delta\;f = f\;\id[\Phi]\;(\den A _\delta\, a)
  .
\]
Pasting allows us to complete the definition of evaluation.
\[
\begin{array}{lcl}
  \teval_\Gamma^A & : & \TmG A \to \den \Gamma \todot \den A \\[1ex]
  \evalG C {\orE t u v} \Delta \; \gamma & = &
    \tpaste^C \;
     (\mapC {(\lambda \Phi\, (\tau  : \Phi \leq \Delta) \to [f,g])}\;c) \\[1ex]
  \multicolumn 1 r {\mbox{ where } c} & = &
      \evalG{A \lor B} t\;\Delta\;\gamma \\
  \multicolumn 1 r {f\;a} & = &
    \eval[\Gamma.A] C u \;\Phi\;(\den\Gamma_\tau\,\gamma, a) \\
  \multicolumn 1 r {g\;b} & = &
    \eval[\Gamma.B] C v \;\Phi\;(\den\Gamma_\tau\,\gamma, b) \\
  \\
\end{array}
\]
\begin{exercise}
  Really complete the definition of $\teval$ (all missing cases).
\end{exercise}
That's it!

\newcommand{\citeentry}[1]{\item\nocite{#1}\bibentry{#1}}

\section{Literature}

Further reading on normalization by evaluation in general:
\begin{enumerate}

\citeentry{bergerSchwichtenberg:lics91}

\citeentry{jungTiuryn:tlca93}

\citeentry{altenkirchHofmannStreicher:ctcs95}

\citeentry{abel:habil}

\end{enumerate}

% \begin{itemize}
% \item \citet{bergerSchwichtenberg:lics91}
% \item \citet{jungTiuryn:tlca93}
% \item \citet{altenkirchHofmannStreicher:ctcs95}
% \item \citet{abel:habil}
% \end{itemize}
\noindent
Literature dealing in particular with the case of disjunction /
disjoint sums:

\begin{enumerate}

\citeentry{danvy:popl96}

\citeentry{fioreSimpson:tlca99}

\citeentry{altenkirchDybjerHofmannScott:lics01}

\citeentry{balatDiCosmoFiore:popl04}

\citeentry{altenkirchUustalu:flops04}

\citeentry{barral:PhD}

\citeentry{scherer:popl17}

\end{enumerate}

% \begin{itemize}
% \item \citet{danvy:popl96}
% \item \citet{fioreSimpson:tlca99}
% \item \citet{altenkirchDybjerHofmannScott:lics01}
% \item \citet{balatDiCosmoFiore:popl04}
% \item \citet{altenkirchUustalu:flops04}

% \item \citet{barral:PhD}
% \item \citet{scherer:popl17}
% \end{itemize}


\section{Conclusion}

\paradot{Acknowledgments}  The presentation of this NbE proof and its
Agda formalization benefitted greately from discussions with Christian
Sattler.  Thanks to Jesper Cockx for listening to a prerun of the talk.

\bibliographystyle{abbrvnat}
%\bibliography{auto-nbeSum}

\end{document}
